---
title: "RDM Lesson"
output: html_notebook
---

# [The racing diffusion model of speeded decision making](https://pubmed.ncbi.nlm.nih.gov/32424622/)

Tillman, G., Van Zandt, T. & Logan, G.D. Sequential sampling models without random between-trial variability: the racing diffusion model of speeded decision making. Psychon Bull Rev 27, 911–936 (2020). <https://doi.org/10.3758/s13423-020-01719-6>

## Simulate a RDM 

#### Config parameters

```{r}
# Initialize variables
a <- 3         # Decision threshold
v1 <- 4        # Drift rate of Accumulator 1
v0 <- 2.84     # Drift rate of Accumulator 0
dt <- 1 / 60   # Sampling rate
```

#### Animation plot of the evidence accumulation process

```{r}
# Required Libraries
library(ggplot2)
library(gganimate)

ntrials <- 1    # Number of trials to include in the GIF

# Initialize Data Frame to Store Simulation Results
simulation_data <- data.frame(
  time = numeric(),
  value = numeric(),
  accumulator = character(),
  trial = integer()
)

# Simulate Trials
set.seed(42)  # For reproducibility
for (trial in 1:ntrials) {
  y1 <- 0  # Starting point for Accumulator 1
  y0 <- 0  # Starting point for Accumulator 2
  x <- 0   # Starting time
  goon <- TRUE
  
  while (goon) {
    # Update Accumulator Values
    y1 <- c(y1, tail(y1, 1) + rnorm(1, mean = v1 * dt, sd = sqrt(dt)))
    y0 <- c(y0, tail(y0, 1) + rnorm(1, mean = v0 * dt, sd = sqrt(dt)))
    x <- c(x, tail(x, 1) + dt)
    
    # Add to Simulation Data
    simulation_data <- rbind(
      simulation_data,
      data.frame(
        time = tail(x, 1),
        value = tail(y1, 1),
        accumulator = "Accumulator 1",
        trial = trial
      ),
      data.frame(
        time = tail(x, 1),
        value = tail(y0, 1),
        accumulator = "Accumulator 2",
        trial = trial
      )
    )
    
    # Check if Boundary is Crossed
    if (tail(y1, 1) > a || tail(y0, 1) > a) {
      goon <- FALSE
    }
  }
}

# Create the Plot
p <- ggplot(simulation_data, aes(x = time, y = value, color = accumulator, group = accumulator)) +
  geom_line(size = 1) +
  geom_hline(yintercept = a, linetype = "dashed", color = "gray") +
  geom_hline(yintercept = 0, linetype = "solid", color = "black") +
  facet_wrap(~trial, ncol = 1, scales = "free_x") +
  theme_minimal() +
  labs(
    title = "Diffusion Model Simulation",
    x = "Time",
    y = "Accumulator Value",
    color = "Accumulator"
  ) +
  theme(
    legend.position = "top",
    plot.title = element_text(size = 16, face = "bold")
  ) +
  transition_reveal(time)

p

```

#### Data simulation

```{r}
# Parameters
ntrials <- 1000   # Number of simulated trials
a <- 3            # Decision threshold
v1 <- 4           # Drift rate of Accumulator 1
v0 <- 2.84        # Drift rate of Accumulator 2
dt <- 1 / 60      # Sampling rate

# Initialize RT (reaction time) and Response (correct or incorrect) vectors
rt <- rep(NA, ntrials)   # Reaction times
resp <- rep(FALSE, ntrials)  # Response: TRUE for Accumulator 1, FALSE for Accumulator 2

# Simulate trials
set.seed(42)  # Set seed for reproducibility

for (idx in 1:ntrials) {
  # Initialize variables for each trial
  x <- 0      # Time starts at 0
  y1 <- 0     # Starting point of Accumulator 1
  y0 <- 0     # Starting point of Accumulator 2
  goon <- TRUE
  
  while (goon) {
    # Simulate the diffusion process
    y1 <- c(y1, tail(y1, 1) + rnorm(1, mean = v1 * dt, sd = sqrt(dt)))
    y0 <- c(y0, tail(y0, 1) + rnorm(1, mean = v0 * dt, sd = sqrt(dt)))
    x <- c(x, tail(x, 1) + dt)
    
    # Check if the boundary is crossed
    if (tail(y1, 1) > a || tail(y0, 1) > a) {
      rt[idx] <- tail(x, 1)             # Record reaction time
      resp[idx] <- tail(y1, 1) > a     # TRUE if Accumulator 1 crosses the threshold
      goon <- FALSE                    # End the trial
    }
  }
}
```

#### RDM function

```{r}
utl_inverse_gaussian_defective <- function(x, drift_pdf, drift_cdf, threshold1, threshold2) {
  # PDF computation
  pdf <- (threshold1) / sqrt(2 * pi * (x^3)) * 
    exp(-0.5 * ((drift_pdf * x - threshold1)^2) / x)
  
  # CDF computation
  cdf <- pnorm(((drift_cdf * x) - threshold2) / sqrt(x)) +
    exp(2 * drift_cdf * threshold2) * 
    pnorm((-(drift_cdf * x) - threshold2) / sqrt(x))
  
  # Defective probabilities
  probs <- pdf * (1 - cdf)
  
  # Return results as a list
  list(g = probs, pdf = pdf, cdf = cdf)
}

```

#### Display RT distributions for the two accumulators

```{r}
# Reaction Time Histograms and Distributions
x_seq <- seq(dt, 3, by = dt)  # Time steps for the distributions

# Accumulator 1
rt1 <- ifelse(resp, rt, -rt)  # Correct responses positive, incorrect negative
dist_acc1 <- utl_inverse_gaussian_defective(x_seq, v1, v0, a, a)

# Plot Accumulator 1
ggplot() +
  geom_histogram(aes(x = rt1, y = ..density..), bins = ntrials / 10, fill = "gray", alpha = 0.5) +
  geom_line(aes(x = x_seq, y = dist_acc1$pdf), color = "blue", size = 1, linetype = "solid") +
  geom_line(aes(x = x_seq, y = 1 - dist_acc1$cdf), color = "red", size = 1, linetype = "solid") +
  geom_line(aes(x = x_seq, y = dist_acc1$g), color = "black", size = 2, linetype = "solid") +
  labs(
    title = "Accumulator 1 - Reaction Time and Distributions",
    x = "Reaction Time (s)",
    y = "Density"
  ) +
  theme_minimal()

# Accumulator 2
rt2 <- ifelse(!resp, rt, -rt)  # Correct responses negative, incorrect positive
dist_acc2 <- utl_inverse_gaussian_defective(x_seq, v0, v1, a, a)

# Plot Accumulator 2
ggplot() +
  geom_histogram(aes(x = rt2, y = ..density..), bins = ntrials / 10, fill = "gray", alpha = 0.5) +
  geom_line(aes(x = x_seq, y = dist_acc2$pdf), color = "blue", size = 1, linetype = "solid") +
  geom_line(aes(x = x_seq, y = 1 - dist_acc2$cdf), color = "red", size = 1, linetype = "solid") +
  geom_line(aes(x = x_seq, y = dist_acc2$g), color = "black", size = 2, linetype = "solid") +
  labs(
    title = "Accumulator 2 - Reaction Time and Distributions",
    x = "Reaction Time (s)",
    y = "Density"
  ) +
  theme_minimal()

# Displaying mean reaction times and accuracies
cat("Results:\n\n")

cat("Correct responses:\n")
cat("Mean RT:", mean(rt[resp], na.rm = TRUE), "\n")
cat("Proportion of correct responses:", mean(resp, na.rm = TRUE), "\n\n")

cat("Incorrect responses:\n")
cat("Mean RT:", mean(rt[!resp], na.rm = TRUE), "\n")
cat("Proportion of errors:", mean(!resp, na.rm = TRUE), "\n")
```

## RDM parameter estimation

```{r}
# Define the initial parameters
x0 <- c(0, 0, 0)  # Initial points

# Define the function to be minimized (negative log-likelihood or fit function)
fit_RDM <- function(x, rt, resp) {
  # Transform parameters to their native space
  a <- exp(x[1])
  v1 <- exp(x[2])
  v0 <- exp(x[3])
  
  # Compute likelihood of responses of Accumulator 1
  P1 <- log(utl_inverse_gaussian_defective(rt[resp == TRUE], v1, v0, a, a)$g)
  
  # Compute likelihood of responses of Accumulator 2
  P2 <- log(utl_inverse_gaussian_defective(rt[resp == FALSE], v0, v1, a, a)$g)
  
  # Sum response likelihood
  loglik <- sum(c(P1, P2))  # Return the total log-likelihood
  
  return(-loglik)
}

# Optimization function for the likelihood function


# Optimization options
options <- list(method = "BFGS")  # Equivalent to 'quasi-newton' in MATLAB

# Run the optimization
result <- optim(
  par = x0, 
  fn = fit_RDM, 
  rt = rt, 
  resp = resp, 
  method = "BFGS",  # Similar to 'quasi-newton' in MATLAB
)

# Extract the results and transform to their native space
p_ddm <- list(
  a = exp(result$par[1]),
  v1 = exp(result$par[2]),
  v0 = exp(result$par[3])
)

# Display results
cat("Results:\n\n")
cat("Parameter estimates:\n")
print(p_ddm)

```
